---
title: "R Notebook"
output: html_notebook
---



```{r}
install.packages("data.table")
library(data.table)
library(foreach)
library(doParallel)
```

# defining functions
```{r}
haversine <- function(lat1, lon1, lat2, lon2) {
  miles_constant <- 3959
  lat1 <- lat1 * pi / 180
  lon1 <- lon1 * pi / 180
  lat2 <- lat2 * pi / 180
  lon2 <- lon2 * pi / 180
  
  dlat <- lat2 - lat1
  dlon <- lon2 - lon1
  
  a <- sin(dlat / 2)^2 + cos(lat1) * cos(lat2) * sin(dlon / 2)^2
  c <- 2 * asin(sqrt(a))
  
  return(miles_constant * c)
}
```


```{r}
# LoadIng  the clinic.csv file (pipe-delimited)
df<-fread("C:/Users/Sreeja/Downloads/clinics.csv")

# Setting reference location
lat_ref <- 40.671
lon_ref <- -73.985
```

#using for-loop approach
```{r}
start_time <- Sys.time()

# Creating  an empty vector for storing distances
df$distance_for <- numeric(nrow(df))

# Computing distances using a for-loop
for (i in 1:nrow(df)) {
  df$distance_for[i] <- haversine(lat_ref, lon_ref, df$locLat[i], df$locLong[i])
}

end_time <- Sys.time()
for_loop_time <- end_time - start_time
print(paste("For-Loop Execution Time:", for_loop_time))
```
# parallel execution using foreach
```{r}
```


```{r}
#set for windows
cl <- makeCluster(detectCores(), type='PSOCK')
registerDoParallel((cl))
```


```{r}
set.seed(42)
```


```{r}
registerDoParallel(cores = detectCores())  # Use all available cores

start_time <- Sys.time()

df$distance_foreach <- foreach(i = 1:nrow(df), .combine = c) %dopar% {
  haversine(lat_ref, lon_ref, df$locLat[i], df$locLong[i])
}

end_time <- Sys.time()
foreach_time <- end_time - start_time
print(paste("Foreach Execution Time:", foreach_time))
```

```{r}
stopCluster(cl)#stopping cluster
```

```{r}
start_time <- Sys.time()

# Computing  distances using a vectorized approach
df[, distance_dt := haversine(lat_ref, lon_ref, locLat, locLong)]

end_time <- Sys.time()
vectorized_time <- end_time - start_time
print(paste("Vectorized Execution Time:", vectorized_time))
```


```{r}
# Creating  a summary table
results <- data.table(
  Method = c("For-Loop", "Foreach (Parallel)", "Vectorized data.table"),
  Execution_Time = c(for_loop_time, foreach_time, vectorized_time)
)

# Printing  results
print(results)
```

#Professor or whoever is grading my assignment,

#My R studio suddenly stopped working while I was workng on the assignment. I tried uninstalling and reinstalling it, but nothing worked. I didnt have any other source, so i completed the assignment using my friends laptop. However, I did not take anyones help in completing the assignment.