---
title: "Week 6 Assignment"
author: "Naga Jahnavi Kancharla"
date: "2025-03-02"
output: html_document
---


```{r}
getwd()
```
```{r}
setwd("C:/Users/Sreeja/Downloads")
```

```{r}
#loading the required libraries
library(data.table)
library(geosphere)
library(GA)
library(purrr)
library(sf)

## Loading the population data
pop.data <- data.table::fread("C:/Users/Sreeja/Downloads/Mo_pop_Sim/Mo_pop_Sim.csv")
str(pop.data)

```
```{r}


# Defining  fraction for sampling
frac <- 0.0001

# Extracting a sample of randomly chosen rows 
small.data <- pop.data[sample(1:nrow(pop.data),
                              size = round(nrow(pop.data) * frac),
                              replace = FALSE), ]

# Loading the FQHC data
data_path <- "C:/Users/Sreeja/Downloads/MO_2018_Federally_Qualified_Health_Center_Locations/MO_2018_Federally_Qualified_Health_Center_Locations/MO_2018_Federally_Qualified_Health_Center_Locations.shp"

# Reading spatial data correctly
fqhc.data <- data.table(as.data.frame(st_read(data_path)))

```

```{r}
#==== APPROACH 1: AVERAGE CLOSENESS ====
# Function to calculate average distance from FQHC to all residences
calculate_avg_distance <- function(fqhc_idx) {
  fqhc_lon <- fqhc.data$Longitude[fqhc_idx]
  fqhc_lat <- fqhc.data$Latitude[fqhc_idx]
  
  distances <- distHaversine(
    matrix(c(fqhc_lon, fqhc_lat), ncol = 2),
    as.matrix(small.data[, .(long, lat)])
  )
  
  # Converting  meters to miles for easier interpretation
  mean(distances) / 1609.34
}

# Calculating  average distances for all FQHCs
fqhc.data$avg_distance_miles <- sapply(1:nrow(fqhc.data), calculate_avg_distance)

# Fitness function for Approach 1 (MINIMIZE average distance)
fitness_function_avg_distance <- function(solution) {
  if(sum(solution) != 8) return(0) # Ensure exactly 8 locations are selected
  
  selected_indices <- which(solution == 1)
  # Lower average distance is better, so we invert it
  return(1 / mean(fqhc.data$avg_distance_miles[selected_indices]))
}

```


```{r}
# APPROACH 2: POPULATION DENSITY
#  calculating  number of residents within 30 miles
calculate_residents_within_radius <- function(fqhc_idx, radius_miles = 30) {
  fqhc_lon <- fqhc.data$Longitude[fqhc_idx]
  fqhc_lat <- fqhc.data$Latitude[fqhc_idx]
  
  distances <- distHaversine(
    matrix(c(fqhc_lon, fqhc_lat), ncol = 2),
    as.matrix(small.data[, .(long, lat)])
  )
  
  radius_meters <- radius_miles * 1609.34
  sum(distances <= radius_meters)
}

# Calculating  residents within 30 miles for all FQHCs
fqhc.data$residents_within_30mi <- sapply(1:nrow(fqhc.data), calculate_residents_within_radius)

# Fitness function for Approach 2 (MAXIMIZE residents served)
fitness_function_density <- function(solution) {
  if(sum(solution) != 8) return(0) # Ensure exactly 8 locations are selected
  
  selected_indices <- which(solution == 1)
  return(sum(fqhc.data$residents_within_30mi[selected_indices]))
}
```


```{r}
# Choosing  which approach to use
use_approach <- 2  # Change to 1 for average closeness approach
fitness_function <- if(use_approach == 1) fitness_function_avg_distance else fitness_function_density

# Runnig  the Genetic Algorithm
ga_result <- ga(
  type = "binary",
  fitness = fitness_function,
  nBits = nrow(fqhc.data),  
  popSize = 50,             
  maxiter = 100,           
  run = 30,                 
  pmutation = 0.1,          
  monitor = TRUE,    
  suggestions = matrix(0, ncol = nrow(fqhc.data), nrow = 1)
)

# Getting  the best solution
best_solution <- ga_result@solution[1, ]
selected_fqhcs <- which(best_solution == 1)

# Printing results
cat("Selected FQHC indices:", selected_fqhcs, "\n")
cat("Selected FQHC names:\n")
for(idx in selected_fqhcs) {
  cat(paste0(idx, ": ", fqhc.data$Facility[idx], " (", 
             fqhc.data$City[idx], ", ", fqhc.data$County[idx], ")\n"))
  
  if(use_approach == 1) {
    cat(paste0("  Average distance: ", round(fqhc.data$avg_distance_miles[idx], 2), " miles\n"))
  } else {
    cat(paste0("  Residents within 30 miles: ", fqhc.data$residents_within_30mi[idx], "\n"))
  }
}
```


I would apply the second strategy and find locations of FQHCs with the highest density of homes situated within thirty miles for this mental health service allocation task. The solution sets out to find locations that will reach the maximum number of potential patients seeking specialized mental health care.
Health services for mental health demand special employees and restricted resources. Placing these critical resources within areas with dense population will increase the number of people who can benefit from these essential services. Specialized mental health services operated at superior levels can expect their target clients to make reasonable journeys to their facilities since immediate access requirements do not apply to this field.
A fitness function must assess the total residents located within thirty miles of every potential FQHC site. The genetic algorithm would optimize this measurement through its operation. The optimization of services is suitable because we aim to reach the maximum number of Missouri residents with our constrained facility capacity (8 facilities). Candidate solutions in the genetic algorithm function as representatives of unique FQHC sets containing 8 facilities which obtain fitness value from adding residents located in a 30-mile range of each chosen area.
Several important assumptions underlie this approach. The analysis starts by considering residents who have no restrictions in traveling distances of 30 miles to reach their mental health support. We have established that resident population numbers adequately represent the size of the mental health service clientele. We believe our chosen population sample correctly represents how the population groups itself throughout all areas. The method assumes that a 30-mile distance around each FQHC facility establishes correct boundaries for mental health service coverage.
A drawback of this methodology leads to the excessive concentration of facilities in dense urban areas which can result in neglected mental health care for rural areas. Among all existing FQHCs we are restricted to select 8 locations so maximizing population coverage becomes the best option for specialized services that need numerous patients to maintain effectiveness and sustain their operations.